Evaluating the Performance and Energy Efficiency of COSMO-ART, a Coupled Numerical Weather Forecast and Chemical Transport Model

### 2014-04-11 paper_enahpc14/Introduction.tex
adds minor changes

COSMO-ART is  computationally much more demanding than  the COSMO core
since a  large number of additional  tracers and processes  have to be
considered.  Thus this model is currently severely-limited in terms of
applicability and expensive in terms of energy consumption.  COSMO has
recently  been  ported  to  GPUs  within the  framework  of  the  High
Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
optimize it  for computational and energy  efficiency.  Although these
developments will  facilitate the  application of COSMO  for numerical
weather prediction and climate  simulations, they do little to address
the  coupling with  the  ART model  extension,  for which  significant
investments  are still required  to take  it to  a similar  level. The
efficiency of ART is being  addressed in the EU Exa2Green project. The
ultimate goal  of the  project is to  deliver a prototype  code, which
provides an energy  efficiency of at least five  times of the baseline
value.   Such   an  implementation   would  allow  the   community  to
investigate critical  questions at  higher resolution and  over longer
periods, at reduced cost  to the environment.

In  Sec.~\ref{sec:1},  we  introduce  the regional  atmospheric  model
COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
chemistry,  aerosols, radiation, and  clouds and  additionally specify
its technical setup related to the investigated performance and energy
evaluation  methods.   Sec.~\ref{sec:2}  describes the  different  HPC
systems   and  associated  power   measurement  equipments   used  for
conducting the benchmarking study and presents energy-aware techniques
employed  to   determine  the  energy  footprint   of  the  considered
application.   Sec.~\ref{sec:3}  gathers  benchmark results  from  the
Exa2Green  project and  highlights  areas where  improvements will  be
necessary for the subsequent  baseline refactoring.  Some related work
will  be  discussed  in  Sec.~\ref{sec:4} Finally,  we  conclude  with
practical  guidelines to identifying  energy saving  opportunities and
give  directions for further  developments of  the modeling  system in
Sec.~\ref{concl}.

>>>>

COSMO-ART is  computationally much more demanding than  the COSMO core
since a  large number of additional  tracers and processes  have to be
considered.  Thus this model is currently severely-limited in terms of
applicability and expensive in terms of energy consumption.  COSMO has
recently  been  ported  to  GPUs  within the  framework  of  the  High
Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
optimize    it    for     computational    and    energy    efficiency
(\url{http://www.hp2c.ch/}).    Although   these   developments   will
facilitate the  application of COSMO for  numerical weather prediction
and climate simulations,  they do little to address  the coupling with
the ART  model extension, for which significant  investments are still
required to take it to a similar level. The efficiency of ART is being
addressed        in       the        EU        Exa2Green       project
(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times of the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

In  Sec.~\ref{sec:1}, we  briefly introduce  the  regional atmospheric
model COSMO-ART \citep{Vogel-2009} that accounts for feedbacks between
chemistry,  aerosols, radiation, and  clouds and  additionally specify
its technical setup related to the investigated performance and energy
evaluation  methods.   Sec.~\ref{sec:2}  describes the  different  HPC
systems   and  associated  power   measurement  equipments   used  for
conducting the benchmarking study and presents energy-aware techniques
employed  to   determine  the  energy  footprint   of  the  considered
application.   Sec.~\ref{sec:3}  gathers  benchmark results  from  the
Exa2Green  project and  highlights  areas where  improvements will  be
necessary for the subsequent  baseline refactoring.  Some related work
will  be  discussed  in  Sec.~\ref{sec:4} Finally,  we  conclude  with
practical  guidelines to identifying  energy saving  opportunities and
give  directions for further  developments of  the modeling  system in
Sec.~\ref{concl}.

### 2014-04-11 paper_enahpc14/Sec2:COSMO-ART.tex
adds minor changes

\subsection{Model Description}
\label{subsec:1.1}
COSMO-ART  (ART  stands  for   Aerosols  and  Reactive  Trace  gases),
developed  at  KIT  Karlsruhe  \citep{Vogel-2009}, is  a  regional  to
continental scale model coupled  online to the COSMO numerical weather
prediction (NWP) and climate model \citep{Baldauf-2011}. It allows the
online calculation of reactive  trace substances and their interaction
with  the atmosphere.   Physical processes  like  transport, turbulent
diffusion,  and  dry and  wet  deposition  are  treated together  with
photochemistry and aerosol dynamics using the modal approach.

Detailed  model  description  can   be  found  in  the  aforementioned
publications  as   well  as  in   \citep{Stanelle-2010,  Bangert-2012,
  Knote-2011, Knote-2013}.

@@

This COSMO-ART  version is configured  to deal with a  semi Lagrangian
horizontal advection scheme  with tricubic interpolation and selective
filling diffusion option in combination with the Runge-Kutta dynamical
core. It  also makes  use of the  Kinetic PreProcessor solver  for the
resolution of  atmospheric chemistry ordinary  differential equations.
Concerning the  modelling of wet deposition in  aerosols, the baseline
has  only  indirect  cloud  feedbacks  but  doesn't  include  in-cloud
scavenging  (rainout)   and  below-cloud  scavenging   (washout)  yet.
Amongst   physical  parameterizations,   precipitation   formation  is
performed  by a  two-moment  cloud microphysics  scheme  instead of  a
classical  bulk microphysics  scheme.

>>>>

\subsection{Model Description}
\label{subsec:1.1}
COSMO-ART  (ART  stands  for   Aerosols  and  Reactive  Trace  gases),
developed  at  KIT  Karlsruhe  \citep{Vogel-2009}, is  a  regional  to
continental scale model coupled  online to the COSMO numerical weather
prediction (NWP) and climate model \citep{Baldauf-2011}. It allows the
online calculation of reactive  trace substances and their interaction
with  the atmosphere.   Physical processes  like  transport, turbulent
diffusion,  and  dry and  wet  deposition  are  treated together  with
photochemistry and aerosol dynamics using the modal approach. Detailed
model description  can be found in the  aforementioned publications as
well    as   in   \citep{Stanelle-2010,    Bangert-2012,   Knote-2011,
  Knote-2013}.

@@

This COSMO-ART version is configured with a semi Lagrangian horizontal
advection  scheme with  tricubic interpolation  and  selective filling
diffusion  option  in  combination   with  the  dynamical  core  using
Runge-Kutta time stepping \citep{}.  It  also makes use of the Kinetic
PreProcessor  solver  \citep{}   for  the  resolution  of  atmospheric
chemistry ordinary  differential equations.  Concerning  the modelling
of wet  deposition in aerosols,  the baseline has only  indirect cloud
feedbacks  but  doesn't  include  in-cloud  scavenging  (rainout)  and
below-cloud    scavenging    (washout)    yet.     Amongst    physical
parameterisations,   precipitation  formation   is   performed  by   a
two-moment  cloud  microphysics scheme  instead  of  a classical  bulk
microphysics scheme.
